The Dynamics of Anthropomorphism in Human-Robot Interaction

### 2014-04-04 paper.tex
refined figure anthropomorphism; new bib item; work on sections 4 and 5

As human-likeness itself is a huge dimension\fxwarning{SL: huge dimension?},
some researchers argue that there might not just be one single type of
anthropomorphism but different levels or gradations. For instance,
\citet{persson_anthropomorphism_2000} proposes to understand anthropomorphism as
a multi-layered phenomenon, namely \emph{primitive categorization},
\emph{primitive psychology}, \emph{folk-psychology}, \emph{traits}, \emph{social
roles}, and \emph{emotional anthropomorphism}. Correspondingly,
\citet{ruijten_introducing_2014} propose to understand anthropomorphism as a
continuum ranging from low to high human likeness, and to measure it as such.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.75\columnwidth]{anthropo.pdf}
    \caption{Anthropomorphism is a psychological phenomenon which arises in a
        social context, during an
    interaction between a human agent and a non-human agent. One can take two main perspectives when trying to explain
    anthropomorphism: a human-centered and an artifact-centered perspective (grey
    triangles). Several external factors influence the anthropomorphism
    phenomenon. Person- and artifact-related ones, but also related to the
context of use as well as the actual time spend interacting with the robot.}

@@

In the following we will only briefly present the artifact-centered perspective to explain anthropomorphism. Then, we provide a more detailed review and synthesis of literature that applies a human-centered perspective to explain anthropomorphism. Concretely, we integrate theories from developmental psychology, cognitive science / neuroscience, and social psychology to shed light on the question why humans anthropomorphize.

@@

\fxwarning{SL: maybe add a ref to previous Julia's paper on anthropomorphism?}

@@

This cognitive viewpoint on anthropomorphism is important because people's
estimation of an agent's knowledge model and its capabilities affects the way
they relate to it. This holds implications for the resulting interaction with
the system, the user experience, and the acceptance. Previous research examined
the validity of the mental model concept with various kinds of robots
\citep{schmitz_concepts_2011,kiesler_mental_2002}. Findings suggest that people
tend to hold richer mental models of anthropomorphic robots in contrast to
mechanic ones \citep{kiesler_mental_2002}\footnote{Later in this article,
we may however come to question how far this result is based on
expectations from the human user side, which are likely to be refined
after continued interaction, and after the user gets acquainted with the
robot's behavior.}. A similar finding is described in
\cite{hegel_understanding_2008} and \cite{krach_can_2008}. In a user study using
functional magnetic resonance imagery (fMRI), the authors found that
participants implicitly attributed human-like qualities, such as mental states,
to their non-human interaction partner. Indeed, from an early age humans develop
a tendency to explain one's own and others' actions in terms of beliefs, desires
and goals,  called \textit{theory of mind}. The theory of mind allows the
implicit attribution of intentions and other mental states to others
\citep{premack1978does,leslie_pretense_1987,Frith2003}.  The finding of
\cite{hegel_understanding_2008} and \cite{krach_can_2008} was evident at the
behavioral level as well as on the neuro-physiological level: the more the
human-likeness of the artefactual partner increases, the more brain areas
associated with theory of mind  get activated \citep{krach_can_2008}.
\footnote{\hl{Ha! This says that anthropomorphism can also occur on a behavioral
level, yes! :) so we should also measure anthropomorphism in the behavior.}}
Implication of human brain area involved in the inference of others` mental
states has also been shown in response to viewing
non-anthropomorphic/non-humanoid robotic gadgets whose behavior has been
described as unpredictable -- but not in response to those whose behavior was
described as predictable -- a finding also found at the behavioral level
\citep{Waytz2010}.  Interestingly, the grey matter volume of a brain area
related to theory of mind has been correlated to individual's score of
anthropomorphism \citep{cullen2013individual}.

At lower, more automatic, level, robots have been shown to elicit resonance
behaviors in the human brain. Resonance behaviors \citep{Rizzolatti1999} are
mechanisms by which the brain areas involved in the internal representations of
an action, an emotion or a sensation are equally recruited during the perception
of another individual performing the same action or experiencing the same
emotion or sensation.  The neurons showing resonant properties have been called
mirror neurons and have been found in a wide range of modalities (motor,
emotional etc..) \fixme{addref}. Seeing a robotic arm reaching for an object
activates the motor mirror neurons the same way as for seeing a human arm
reaching for the same object \citep{Gazzola2007, oberman_eeg_2007}. However, a
study of emotion perception on a robotic face found reduced activity in
emotional brain area known to have mirror properties in response to the robotic
face compared to a human face \citep{Chaminade2010}. Nevertheless, these authors
showed that when the participants were explicitly instructed to pay attention to
the robot's emotional expression, motor mirror neurons get activated.
Interestingly, using electroencephalography,  it has also been shown that
emotional behavior elicited by a robot face influences the speed of human
responses as well as early brain processes the same way as emotional expression
of a human face did \citep{Dubal2010}.

@@

\begin{enumerate}
\item ~the characteristics of the \textbf{robot's design} (degree of anthropomorphic form),
\item ~the psychological determinants in the \textbf{human user} (see
\cite{epley_seeing_2007}),
\item ~the characteristics of the \textbf{context / situation} in which the interaction occurs, and
\item ~the fact of the user getting familiar with the robot \textbf{over time} (duration of interaction)
\end{enumerate}

@@

In the introduction, we reported on a few selected experimental results that
illustrate how the user tendency to anthropomorphize decreased over time.
Combined with the extensive discussion on the nature and factors of
anthropomorphism that we proposed at section~\ref{sec:anthropomorphism}, we
introduce in the following section a novel qualitative model of the evolution
of anthropomorphism over time, \ie a model of the \emph{dynamics of anthropomorphism}.

@@

The actual maximum value of anthropomorphic effects depends on each unique
combination of a human, a robot, a situation and several other factors we introduce below. Thus, the maximum value varies according to the specific characteristics of these factors. Figure~\ref{fig:dynamics} pictures this maximum value at
$t_{max}$, which actually corresponds to what we call the novelty effect peak.\footnote{We assume that there is a peak in the human's experiences of the novelty effects of the unfamiliar system. Further, we suspect that this novelty peak comes along with the maximum value of anthropomorphic effects.} 
 The general
\emph{shape} of the model remains however the same and depicts the evolution of
anthropomorphism over time, \ie the general dynamics of anthropomorphism.\fixme{Last sentence: Do you mean the shape of the curve? i would not agree that it remains the same ... but say that it shows a general tendency to decrease, however that there might be variations.}

@@

\fxwarning{SL: Mention the pre-interaction phase}

@@

@@

In this context, \emph{familiarization} means \emph{to get acquainted with}.
It is not to be confused with the \emph{familiarity thesis} that relates to the
projection of already known cognitive models onto the robot. We discuss this other
aspect at section~\ref{sec:cognition-neuroscience}.

@@

We also discuss two specific aspects: the cognitive impact of disruptive
behaviors (section~\ref{sec:disruptive}) on mutual modelling, and how our models
relate to the \emph{in-the-moment} versus \emph{reflective} perspectives
introduced by~\citet{takayama_perspectives_2012}.

@@

We propose three different cognitive stages (Figure~\ref{fig:cognitivemodel}),
which are related but do not directly match the three phases of
the model of the dynamics of anthropomorphism.

@@

@@

This mental process depends on the human understanding of the robot's
inner working, as well as his/her own tendency to anthropomorphize (the
\emph{personality} in ICA factor), but at this
stage, the \emph{perception} of the robot (its shape for instance) and its
intended \emph{purpose} play a less important role. It is mostly a human-centric
process.  The result of this third phase would be an iteratively adapted
cognitive model of the robot. It is also at this last stage that the human might
modified its own abilities to anthropomorphize, having built a full cognitive
model of the robot and now being able to attribute internal state to the robot
through his theory of mind\fxwarning{SL: last part not so clear. Rephrase?}.

\fxwarning{SL: mention that the stages are not mutually exclusive and actually
overlap over time.}

@@

Common observation of naive people (children or adults) interacting with robots
shows that unexpected behaviors of the robot can have a notable impact on
interaction. \hl{cite literature, e.g. cheating robot, disagreeing, disobeying,
HRI2013 paper}
This is supported by the results from~\citet{Waytz2010}, mentioned in section
\ref{sec:explanations}, that show that people attribute more easily
anthropomorphic features to artefacts when they have unpredictable behaviors.

@@

We emit the hypothesis that an unexpected robot behavior might lead the user to
suddenly restate his/her behavioral model of the robot and will temporarily lead
to an increase or decrease of the level of anthropomorphism (depicted by the
spikes on Figure~\ref{fig:dynamics}).

@@

If the unexpected behavior is not planned, and perceived as such by the human
(case I), the human can interpret that the robot is able to fail. If the robot
explicitly states its failure (for instance, by saying ``I'm lost!''), the
behavior is then called \emph{transparent} \fixme{cite here, Julia?}
\hl{(literature on transparency)}, and the user may hypothesize that the
robot has \emph{introspective} capabilities (it can reflect on its own internal
state), which may lead to higher anthropomorphic projections.  On the
contrary, if the robot shows no sign of recognizing its own failure, the user
may ascribe a lower level of anthropomorphism to the robot.

@@

Case IIIa and IIIb correspond to misinterpretations of the robot behavior. Case
IIIb may actually lead to an increased level of anthropomorphism since the human
will (wrongfully) attribute intentionality to the robot, while case IIIa is
expected to lead to a lower anthropomorphism level.  However, in those two
cases, the next occurrence of an expected behavior, if correctly interpreted
(case I or II), is likely to lead to stronger effects due to a larger delta
between expectations and actual observed behavior. \hl{(we actually assume that
anthropomorphism can be radically changing and be a very spontaneous
"reaction".)}

@@

\subsection{\emph{In-the-moment} versus \emph{Reflective} perspectives}

\fxwarning{SL: Bridge + integration in the section 4 TDB}

Perception of agency can be treated as anthropomorphism.
\citep{takayama_perspectives_2012} proposes to differentiate between a
\textit{in-the-moment} and \textit{reflective} perspective to better understand
how human users perceive agency in robots.\footnote{``The other important
distinction between the two perspectives is the \hl{cognitive processes} that
they engage.''} According to Takayama, an \textit{in-the-moment} perspective
refers to one's most immediate (sometimes visceral) sense in a situation. In
contrast, a \textit{reflective} perspective refers to one's sense of a situation
upon more distanced cogitation and consideration. For instance, we reflectively
might not perceive agency in a social robot, but it can feel quite differently
in the moment of interaction. A critical dimension of agency is the perspective
from which something seems to have agency.  \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive.  That means, in the initial phase of interacting with a
novel device \textbf{(initialization)} people might not consciously respond but
rather mindlessly \citep{nass_machines_2000}. Only after some time of
\textbf{familiarization}, they might respond in a more reflective manner. This
can be illustrated by the fact that people tend to deny interacting with
computational systems as if they were people and yet they respond to computers
in many ways that are remarkably similar to how they respond to people
\citep{reeves_media_1996}.  \cite{takayama_perspectives_2012} also applies a
cognitive viewpoint on the two different perspectives of in-the-moment \vs
reflective to illustrate differences. She states that in-the-moment perceptions
of agency are largely shaped by bottom-up perceptual processes, evoking very
immediate responses (\eg to the cues emitted by the artifact's design, see
Section \ref{sec:anthropomorphic-design}).  In contrast, according to Takayama,
reflective perceptions are more often shaped by top-down processes because of
the nature of reflective thought.

>>>>

As human-likeness itself is manifold and complex,
some researchers argue that there might not just be one single type of
anthropomorphism but different levels or gradations. For instance,
\citet{persson_anthropomorphism_2000} proposes to understand anthropomorphism as
a multi-layered phenomenon, namely \emph{primitive categorization},
\emph{primitive psychology}, \emph{folk-psychology}, \emph{traits}, \emph{social
roles}, and \emph{emotional anthropomorphism}. Correspondingly,
\citet{ruijten_introducing_2014} propose to understand anthropomorphism as a
continuum ranging from low to high human likeness, and to measure it as such.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.75\columnwidth]{anthropo.pdf}
    \caption{Anthropomorphism is a psychological phenomenon which arises in a
        social context, during an
    interaction between a human agent and a non-human agent. One can take two main perspectives when trying to explain
    anthropomorphism: a human-centered and an artifact-centered perspective (grey
    triangles). Several external factors influence the anthropomorphism
    phenomenon. Person- and artifact-related characteristics, but also related to the
context of use as well as the actual time spent interacting with the robot (interaction history).}

@@

In the following we will only briefly present the artifact-centered perspective to explain anthropomorphism\footnote{A review of anthropomorphic design of robots can also be found in \cite{fink_anthropomorphism_2012}}. Then, we provide a more detailed review and synthesis of literature that applies a human-centered perspective to explain anthropomorphism. Concretely, we integrate theories from developmental psychology, cognitive science / neuroscience, and social psychology to shed light on the question why humans anthropomorphize.

@@

@@

This cognitive viewpoint on anthropomorphism is important because people's
estimation of an agent's knowledge model and its capabilities affects the way
they relate to it. This holds implications for the resulting interaction with
the system, the user experience, and the acceptance. Previous research examined
the validity of the mental model concept with various kinds of robots
\citep{schmitz_concepts_2011,kiesler_mental_2002}. Findings suggest that people
tend to hold richer mental models of anthropomorphic robots in contrast to
mechanic ones \citep{kiesler_mental_2002}\footnote{Later in this article,
we may however come to question how far this result is based on
expectations from the human user side, which are likely to be refined
after continued interaction, and after the user gets acquainted with the
robot's behavior.}. A similar finding is described in
\cite{hegel_understanding_2008} and \cite{krach_can_2008}. In a user study using
functional magnetic resonance imagery (fMRI), the authors found that
participants implicitly attributed human-like qualities, such as mental states,
to their non-human interaction partner. Indeed, from an early age humans develop
a tendency to explain one's own and others' actions in terms of beliefs, desires
and goals,  called \textit{theory of mind}. The theory of mind allows the
implicit attribution of intentions and other mental states to others
\citep{premack1978does,leslie_pretense_1987,Frith2003}.  The finding of
\cite{hegel_understanding_2008} and \cite{krach_can_2008} was evident at the
behavioral level as well as on the neuro-physiological level: the more the
human-likeness of the artefactual partner increases, the more brain areas
associated with theory of mind  get activated \citep{krach_can_2008}.
Implication of human brain area involved in the inference of others` mental
states has also been shown in response to viewing
non-anthropomorphic/non-humanoid robotic gadgets whose behavior has been
described as unpredictable -- but not in response to those whose behavior was
described as predictable -- a finding also found at the behavioral level
\citep{Waytz2010}.  Interestingly, the grey matter volume of a brain area
related to theory of mind has been correlated to individual's score of
anthropomorphism \citep{cullen2013individual}.

At lower, more automatic, level, robots have been shown to elicit resonance
behaviors in the human brain. Resonance behaviors \citep{Rizzolatti1999} are
mechanisms by which the brain areas involved in the internal representations of
an action, an emotion or a sensation are equally recruited during the perception
of another individual performing the same action or experiencing the same
emotion or sensation.  The neurons showing resonant properties have been called
mirror neurons and have been found in a wide range of modalities (motor,
emotional etc.) \fixme{addref}. Seeing a robotic arm reaching for an object
activates the motor mirror neurons the same way as for seeing a human arm
reaching for the same object \citep{Gazzola2007, oberman_eeg_2007}. However, a
study of emotion perception on a robotic face found reduced activity in
emotional brain area known to have mirror properties in response to the robotic
face compared to a human face \citep{Chaminade2010}. Nevertheless, these authors
showed that when the participants were explicitly instructed to pay attention to
the robot's emotional expression, motor mirror neurons get activated.
Interestingly, using electroencephalography,  it has also been shown that
emotional behavior elicited by a robot face influences the speed of human
responses as well as early brain processes the same way as emotional expression
of a human face did \citep{Dubal2010}.

@@

\begin{enumerate}
\item ~the characteristics of the \textbf{robot's design} (degree of anthropomorphic form),
\item ~the psychological determinants in the \textbf{human user} (see
\cite{epley_seeing_2007}),
\item ~the characteristics of the \textbf{context / situation} in which the interaction occurs, and
\item ~the fact of the user getting familiar with the robot \textbf{over time} (duration / history of interaction)
\end{enumerate}

@@

In the introduction, we reported on a few selected experimental results that
illustrate how the user tendency to anthropomorphize decreased over time.
Combined with the extensive discussion on the nature and factors of
anthropomorphism that we proposed in section~\ref{sec:anthropomorphism}, we
introduce in the following section a novel qualitative model of the evolution
of anthropomorphism over time, \ie a model of the \emph{dynamics of anthropomorphism}.

@@

The actual maximum value of anthropomorphic effects depends on each unique
combination of a human, a robot, a situation and several other factors we introduce below. Thus, the maximum value varies according to the specific characteristics of these factors. Figure~\ref{fig:dynamics} pictures this maximum value at
$t_{max}$, which actually corresponds to what we call the novelty effect peak.\footnote{We assume that there is a peak in the human's experiences of the novelty effects of the unfamiliar system. Further, we suspect that this novelty peak comes along with the maximum value of anthropomorphic effects.} 
 The general
\emph{shape} of the model remains however the same and depicts the evolution of
anthropomorphism over time, \ie the general dynamics of anthropomorphism.\fixme{JF: Last sentence: Do you mean the shape of the curve? I would not agree that it remains the same ... but say that it shows a general tendency to decrease, however that there might be variations, mostly determined by the 4 factors.}

@@

@@

\paragraph{Pre-interaction phase\\}
\hl{needs to be written, \eg people form expectations \etc and this leads to the ICA (\ICA).}

@@

In this context, \emph{familiarization} means \emph{to get acquainted with}.
It is not to be confused with the \emph{familiarity thesis} that relates to the
projection of already known cognitive models onto the robot. We discuss this other
aspect in section~\ref{sec:cognition-neuroscience}.

@@

We also discuss two specific aspects: the cognitive impact of disruptive
behaviors (section~\ref{sec:disruptive}) on mutual modelling, and how our models
relate to the \emph{in-the-moment} versus \emph{reflective} perspectives on agency in HRI~\citet{takayama_perspectives_2012}.

@@

We propose three different cognitive stages (Figure~\ref{fig:cognitivemodel}),
which are related but do not directly match the three phases of
the model of the dynamics of anthropomorphism. Note that these three stages are not mutually exclusive, they may overlap in time.

@@

The intuitive anthropomorphization in the cognitive stage 1 might be characteristic of what \cite{takayama_perspectives_2012} calls the \textit{in-the-moment} perspective on agency with a robot compared to the \textit{reflective} perspective (see paragraph ``artifact-centered perspective'' in Section \ref{sec:anthropomorphism}). 
For instance, we reflectively
might \textit{not} perceive agency in a social robot, but it can feel quite differently
in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. That means, in the initial phase of interacting with a
novel device \textit{(initialization)} (see Figure \ref{fig:dynamics}) people might not respond consciously but
rather mindlessly \citep{nass_machines_2000}, and in turn anthropomorphize more.
Only after some time of
\textit{familiarization}, they might respond in a more reflective manner (in turn, the tendency to anthropomorphize might decrease). This
can be illustrated by the fact that people tend to deny interacting with
computational systems as if they were people and yet they respond to computers
in many ways that are remarkably similar to how they respond to people
\citep{reeves_media_1996}. \cite{takayama_perspectives_2012} also applies a
cognitive viewpoint on the two different perspectives of in-the-moment \vs
reflective to illustrate differences. She states that in-the-moment perceptions
of agency are largely shaped by bottom-up perceptual processes, evoking very
immediate responses (\eg to the cues emitted by the artifact's design (``artifact-centered perspective''). In contrast, according to Takayama,
reflective perceptions are more often shaped by top-down processes because of
the nature of reflective thought.

@@

This mental process depends on the human understanding of the robot's
inner working, as well as his/her own tendency to anthropomorphize (the
\emph{personality} in ICA factor), but at this
stage, the \emph{perception} of the robot (its shape for instance) and its
intended \emph{purpose} play a less important role. It is mostly a human-centric
process.  The result of this third phase would be an iteratively adapted
cognitive model of the robot. It is also at this last stage that the human might have a
modified tendency to anthropomorphize, because he/she has built a full cognitive
model of the robot. Consequently, the human would now be able to attribute appropriate internal states to the robot
through his/her theory of mind.

@@

Common observation of naive people (children or adults) interacting with robots
shows that unexpected behaviors of the robot can have a notable impact on
interaction. This is supported by the results from~\citet{Waytz2010}, mentioned in section
\ref{sec:cognitive-expl}, that show that people attribute more easily
anthropomorphic features to artifacts when they have unpredictable behaviors.
Also \cite{short_no_2010} found that variations in a robot's behavior (such that it is unexpected) influence people's attribution of mental states and intentionality to a robot. Participants who were playing the ``rock-paper-scissors'' game with a robot that tried to cheat (either verbally or in its action), displayed a greater level of social engagement and made greater attributions of mental states.
However, an unexpected (disruptive) robot behavior might also have a negative effect on user's engagement with the robot. For instance, when the user interprets the robot's behavior as a failure, he/she might lose trust in the robot. \cite{desai_impact_2013} found that in an interaction scenario, especially early failures of the robot can cause a dramatically lower real-time trust, compared to middle or late drops of reliability of the robot.

@@

We emit the hypothesis that an unexpected robot behavior might lead the user to
suddenly restate his/her behavioral model of the robot and will temporarily lead
to an increase or decrease of the level of anthropomorphism (depicted by the
spikes in Figure~\ref{fig:dynamics}).

@@

If the unexpected behavior is not planned, and perceived as such by the human
(case I), the human can interpret that the robot is able to fail. If the robot
explicitly states its failure (for instance, by saying ``I'm lost!''), the
behavior is then called \emph{transparent} \citep{kim_who_2006}, and the user may hypothesize that the
robot has \emph{introspective} capabilities (it can reflect on its own internal
state), which may lead to higher anthropomorphic projections.  On the
contrary, if the robot shows no sign of recognizing its own failure, the user
may ascribe a lower level of anthropomorphism to the robot.

@@

Case IIIa and IIIb correspond to misinterpretations of the robot behavior. Case
IIIb may actually lead to an increased level of anthropomorphism since the human
will (wrongfully) attribute intentionality to the robot, while case IIIa is
expected to lead to a lower anthropomorphism level.  However, in those two
cases, the next occurrence of an expected behavior, if correctly interpreted
(case I or II), is likely to lead to stronger effects due to a larger delta
between expectations and actual observed behavior.

@@

%\subsection{\emph{In-the-moment} versus \emph{Reflective} perspectives}
%
%\fxwarning{SL: Bridge + integration in the section 4 TDB}
%
%\hl{(we actually assume that
%anthropomorphism can be radically changing and be a very spontaneous
%"reaction".)}
